---
title: "GIS713-Homework2"
author: "Aaron_Siegle"
date: "2026-09-15"
output: html_document
---

```{r}
# download packages 
library(RColorBrewer)
library(spatstat)
library(sf)
library(raster)
library(data.table)
library(classInt)
```

```{r}
# Path to data 
path_to_data <- "C:/Users/aasie/OneDrive/Documentos/GIS713-Homework2"
```


```{r}
# Import data
paracou_shp <- shapefile(file.path(path_to_data, "paracou.shp"))
paracou_window <- shapefile(file.path(path_to_data, "paracou_window.shp"))
```

```{r}
names(paracou_shp)
names(paracou_window)
```
```{r}
# Calculate diameter at breast height from BasalArea 
paracou_shp$DBH <- sqrt((paracou_shp$BasalArea/pi) *2)
print(paracou_shp)
```


```{r}
#1: Create a map of the tree locations with symbol sizes proportional to diameter at breast height, and which distinguishes V. Americana and Q. Rosea from each other and other species.
# Make the DBH values continuous
jenks_dbh <- classIntervals(paracou_shp$DBH, n=5, style= "jenks")
dbh_breaks <- jenks_dbh$brks
dbh_vals <- paracou_shp$DBH
cex_vals <- (dbh_vals - min(dbh_vals)) / (max(dbh_vals) - min(dbh_vals)) * 3 + 0.5
# Make the plot
point_col <- adjustcolor("black", alpha.f = 0.5)
plot(paracou_shp,
     pch = as.integer(as.factor(paracou_shp$Species)),
     cex = cex_vals,
     col = point_col)
title("Diameter at Breast Height and Species of Trees in the Paracou Forest")
legend("topleft", 
       legend = unique(paracou_shp$Species), 
       pch = unique(as.integer(as.factor(paracou_shp$Species))),
       title = "Species")
# Making the continuous color palette and breaks
cex_breaks <- (dbh_breaks - min(dbh_vals)) / (max(dbh_vals) - min(dbh_vals)) * 3 + 0.5
legend("topright", 
       legend = round(dbh_breaks, 1),
       pch = 21,
       pt.cex = cex_breaks, 
       title = "Diameter at\nBreast Height (cm)")
```

```{r}
# Question 2: Is there evidence for deviation from CSR expectation for all trees, V. Americana alone, or Q. Rosea alone in nearest neighbor distances for all trees? Provide plots and text supporting your findings (making sure to compute simulation envelopes if necessary).
# Convert the paracou data from a sf object to a ppp object 
paracou_win <- as.owin(st_as_sf(paracou_window))
paracou_ppp <- ppp(
  x = coordinates(paracou_shp)[, 1], 
  y = coordinates(paracou_shp)[, 2], 
  window = paracou_win, 
  marks = paracou_shp$Species
)
```


```{r}
# Testing for V. Americana 
# Subset data
va_ppp <- paracou_ppp[marks(paracou_ppp) == "V. Americana"]

# CSR Test
set.seed(42)
ll_env_va <- envelope(va_ppp, Gest)

# Plot results
plot(ll_env_va, 
    main = "G Function for V. Americana",
    xlab = "r", 
    ylab = "G(r)"
    )                            
```
The expected distribution of V. Americana trees falls within the credibility interval for most distances but then falls outside the credibility interval between 7-8 meters and at 11 meters.At these distances, the distribution of V. Americana exceeds the upper bound of the credibility intervals by small amounts but does not constitute a significant deviation away from CSR. As a whole, we can say that V. Americana does not deviate away from CSR. 

```{r}
# Testing for Q. Rosea 
# Subset data
qr_ppp <- paracou_ppp[marks(paracou_ppp) == "Q. Rosea"]

# CSR Test
set.seed(42)
ll_env_qr <- envelope(qr_ppp, Gest)

# Plot results
plot(ll_env_qr, 
    main = "G Function for Q. Rosea",
    xlab = "r", 
    ylab = "G(r)"
    )      

```
The distribution of Q. Rosea trees falls within the credibility interval for all nearest neighbor distances within a 35 meter radius. The distribution resembles a step function pattern that oscillates around the G-Function while remaining within the boundaries of the credibility interval. We fail to reject the null hypothesis that the Q. Rosea distribution resembles CSR.

```{r}
# CSR Test for all trees 
set.seed(42)
ll_env_all <- envelope(paracou_ppp, Gest)

# Plot results
plot(ll_env_all, 
    main = "G Function for All Trees in the Paracou Forest",
    xlab = "r", 
    ylab = "G(r)"
    )      
```
The observed distribution of trees in the Paracou forest closely follows the G Function but still exceeds the credibility intervals at numerous distances. The credibility interval is smaller here than before because we have a larger number of events, which in this case are trees, to test against CSR. Despite these slight deviations, the distribution of trees largely resembles CSR as a whole. 

```{r}
# Question 3: Are there fewer/more neighboring trees at particular radiuses for all trees, V. Americana, and Q. Rosea than would be expected under CSR? Provide plots and text supporting your findings (making sure to compute simulation envelopes if necessary). What does this mean in the "real world"?
# Use F function here 
# Testing for V. Americana
set.seed(42)
ll_env_va_f <- envelope(va_ppp, Fest)

# Plot results
plot(ll_env_va_f, 
    main = "F Function for V. Americana",
    xlab = "r", 
    ylab = "G(r)"
    )             
```
There are fewer neighboring V. Americana than expected under CSR at distances greater than 15 meters and similar number of neighboring V. Americana as would be expected under CSR at distances nearer than 15 meters. In the world, this means that you are more likely to see other V. Americana around the original tree but as you travel further away from the V. Americana tree you are less likely to see other V. Americana trees. 
```{r}
# Testing for Q. Rosea 
set.seed(42)
ll_env_qr_f <- envelope(qr_ppp, Fest)

# Plot results
plot(ll_env_qr_f, 
    main = "F Function for Q. Rosea",
    xlab = "r", 
    ylab = "G(r)"
    )            
```
You are as likely to see a neighboring Q. Rosea across different spatial scales. You could say that the distribution of Q. Rosea trees is independent from the distance of other Q. Rosea trees. In this way, the dispersion of Q. Rosea trees is random distributed. In the real world, this means that you are no more likely to see a Q. Rosea tree 1 meter away from another Q. Rosea tree than you are to encounter one 30 meters away from the same Q. Rosea tree. 

```{r}
# Testing for all trees 
set.seed(42)
ll_env_all_f <- envelope(paracou_ppp, Fest)

# Plot results
plot(ll_env_all_f, 
    main = "F Function for All Trees in the Paracou Forest",
    xlab = "r", 
    ylab = "G(r)"
    )            
```
When you look at all the trees in the Paracou forest, the amount of neighboring trees matches the expectations for CSR when looking under distances of 2.5 meters. When you look at distances greater than 2.5 meters, there more neighboring trees than you would expect under strict CSR.However, this deviation from the expected value is not significant as the distribution of trees hugs the upper boundary of the credibility interval while still following the overall trend of the expected distribution. In the real world, this means that you are more likely to see trees that are further than 2.5 meters apart than you are likely to see trees that are directly adjacent to each other. 

```{r}
# Question 4: Is there evidence of attraction/repulsion between V. Americana and Q. Rosea species? Again, provide any plots (with simulation envelopes where necessary) to support your assertions.
# function to automate K to L conversion 
KtoL <- function(Kobj){
    # Function to convert Kest objects to Lest objects
    list(
        r=Kobj$r,
        obs=sqrt(Kobj$obs / pi) - Kobj$r,
        lo=sqrt(Kobj$lo / pi) - Kobj$r,
        hi=sqrt(Kobj$hi / pi) - Kobj$r,
        theo=sqrt(Kobj$theo / pi) - Kobj$r
    )
}
```

```{r}
# plotting function for Kest envelopes 
PlotKenvAsL <- function(Kobj, poly_col="#BDD7E7", obs_col="#636363", exp_col="#A50F15", ...){
    Lobj <- KtoL(Kobj)
    polyx <- c(Lobj$r, rev(Lobj$r))
    polyy <- c(Lobj$lo, rev(Lobj$hi))
    xlim <- range(Lobj$r)
    ylim <- range(c(Lobj$obs, Lobj$theo, Lobj$lo, Lobj$hi))
    plot(NA, NA, xlim=xlim, ylim=ylim, type="n", xlab=expression(italic("r")), ylab=expression(italic("L(r)")), ...)
    polygon(x=polyx, y=polyy, col=poly_col, border=NA)
    points(Lobj$r, Lobj$obs, type="l", lty=1, col=obs_col, lwd=2)
    if(length(Lobj$theo) > 0) points(Lobj$r, Lobj$theo, type="l", lty=2, col=exp_col)
}
```


```{r}
# Creating my own shift function because RShift is not working due to the irregular shape 
MyShift <- function(x, max_shift = 50) {
  original_marks <- marks(x)
  shifted <- rshift(x, 
                    group = as.factor(marks(x) == "V. Americana"),
                    edge = "erode")
  marks(shifted) <- original_marks
  return(shifted)
}
```


```{r}
# Using Ripley's K to test for attraction/repulsion 
fJ <- function(x) marks(x) == "V. Americana"
fA <- function(x) marks(x) == "Q. Rosea"

ll_env_torodial <- envelope(paracou_ppp, 
                            Kmulti,
                            I=fJ,
                            J=fA,
                            simulate = expression(
                              MyShift(paracou_ppp)))
PlotKenvAsL(ll_env_torodial)
```

```{r}
# Adding DBH to paracou_ppp
marks(paracou_ppp) <- paracou_shp$DBH
```


```{r}
# Question 5: Assume that juvenile trees are those with DBH <= 14 and mature trees have DBH >= 40. Is there evidence for a nonrandom spatial process producing the observed pattern of juvenile and mature trees? What might this mean in terms of ecology? As before: provide any plots (with simulation envelopes where necessary) to support your assertions.
MyShift_Five <- function(x){
  x_a <- x[marks(x) <= 14]
    x_b <- x[marks(x) >= 40]
    x_a <- rshift(x_a, edge="erode")
    x_b <- rshift(x_b, edge="erode")
    X <- superimpose(x_a, x_b, W=paracou_win)
    return(X)
}
```


```{r}
fJ <- function(x) marks(x) <= 14
fA <- function(x) marks(x) >= 40

ll_env_toroidal_mat <- envelope(paracou_ppp, Kmulti, I=fJ, J=fA,
                                simulate=expression(MyShift_Five(paracou_ppp)))
PlotKenvAsL(ll_env_toroidal_mat)
```

```{r}
str(paracou_shp$DBH)          # should be numeric
sum(paracou_shp$DBH <= 14, na.rm=TRUE)  # juvenile count > 0?
sum(paracou_shp$DBH >= 90, na.rm=TRUE)  # mature count > 0?
```
```{r}
summary(paracou_shp$DBH)
hist(paracou_shp$DBH, breaks=50)
sum(is.na(paracou_shp$DBH))
```

